### 2019-03-03 docs/notebooks/results.Rmd
change abstract back in results.ipynb; somehow overwrote last edit

## Abstract
Convolutional neural networks (CNNs) provide cutting edge tools for computer vision tasks like image classification and scene segmentation. Many mechanisms proposed to improve the accuracy of CNNs been given the name "attention", although it is becoming clear that accuracy alone does not explain how CNNs perform in real-world situations. Attention has a long history of being studied by cognitive scientists and neuroscientists in the context of vision. The CNNs developed in the past few years are also increasingly being used as models of the visual system in the brain. Here we study whether models of attention in the brain, and visual search tasks used to investigate attention, provide an alternate way of benchmarking existing CNNs. We are pursuing these alternate benchmarks in the course of our investigations of attention-like mechanisms for continual learning algorithms. We adapt a classic form of visual search task so it can be performed by neural networks, and measure the accuracy of two CNN architectures, AlexNet and VGG16, as they report whether a target is present among distractors. Surprisingly, we find that both AlexNet and VGG16 show attention-like effects when performing the visual search task. When the target resembles the distractors, then a network's accuracy drops as the number of distractors increases. This dependency of accuracy on the number of items is often referred to as "set size effects". We do not find that these set size effects can be explained by known factors to affect CNNs, such as edge effects. We then adapt an analysis of visual search tasks based on signal detection theory to show that set size effects persist even when controlling for how discriminable the targets are from the distractors. This analysis suggests the "attentional" effects can be attributed to the decision criteria used by the network. We consider how this type of analysis can be extended to images besides visual search stimuli. Then we discuss the implications of our results for comparing CNN output with human behavior, and measuring how any proposed attentional mechanisms change neural network outputs.

@@

>>>>

## Abstract
Convolutional neural networks (CNNs) provide cutting edge tools for computer vision tasks like image classification and scene segmentation. Many mechanisms described as "attention" have been proposed to improve the accuracy of CNNs, although it is becoming clear that accuracy alone does not explain how CNNs perform in real-world situations. Attention has a long history of being studied by cognitive scientists and neuroscientists in the context of vision. The CNNs developed in the past few years are also increasingly being used as models of the visual system in the brain. Here we study whether models of attention in the brain, and visual search tasks used to investigate attention, provide an alternate way of benchmarking existing CNNs. We are pursuing these alternate benchmarks in the course of our investigations of attention-like mechanisms for continual learning algorithms. We adapt a classic form of visual search task so it can be performed by neural networks, and measure the accuracy of two CNN architectures, AlexNet and VGG16, as they report whether a target is present among distractors. Surprisingly, we find that both AlexNet and VGG16 show attention-like effects when performing the visual search task: when the target resembles the distractors, then a network's accuracy drops as the number of distractors increases. This dependency of accuracy on the number of items is often referred to as "set size effects". We do not find that these set size effects can be explained by factors known to affect CNNs, such as edge effects. We then adapt an analysis of visual search tasks based on signal detection theory to show that set size effects persist even when controlling for how discriminable the targets are from the distractors. This analysis suggests the "attentional" effects can be attributed to the decision criteria used by the network. We consider how this type of analysis can be extended to images besides visual search stimuli. Then we discuss the implications of our results for comparing CNN output with human behavior, and measuring how any proposed attentional mechanisms change neural network outputs.

@@

### A psychophysics analysis of visual search shows set size affects remain even when controlling for how discriminable targets are from distractors
The Feature Integration Theory of visual search attributed set size effects to serial processing (Treisman). This conclusion made sense based on the result that reaction time increased when discriminating targets from distractors required considering more than one feature, e.g. both color and orientation. However, later work showed that reaction time effects disappear as targets and distractors become more easily discriminable along one feature dimension (Duncan paper).  
Later researchers, using methods from psychophysics and signal-detection theory, carefully controlled for the discriminability of targets and distractors (Palmer review). This work provided important alternative explanations for the results seen in visual search experiments: attentional effects can be explained in terms of completely parallel but noisy representations that are used by a decision rule (Palmer, Eckstein). This approach also explains errors in terms of low thresholds used by decision making criteria.  We adapted these psychophysics methods to see if they offered additional insights into the attention-like effects we saw in CNNs.  
#### Psychometric curves
First we generated 'feature' search stimuli--red vertical rectangle v. green vertical rectangle--where the red and green were more or less discriminable. We did so by taking advantage of RGB color space: by adding more red to the green color, and vice versa, the colors of the target and distractor can be moved closer to each other from the two red and green verticies of the triangle that defines RGB space. (When the color has "all red" and "all green", it becomes yellow, so the stimuli we generated moved closer and closer to yellow.) We then used these stimuli to measure the accuracy of the neural nets we trained as the target and distractor became more and more similar.  
<img src="../static/psychometric_function_fits.png" alt="efficient" style="width: 500px;"/>
#### Threshold v. set size
We then used the psychometric curves to find what level of target-distractor discriminability gave rise to 75% accuracy, known as the "difference threshold". We found this difference threshold for each trained network and each set size.  
<img src="../static/threshold_v_setsize.png" alt="efficient" style="width: 600px;"/>  
Here we see that the threshold goes up with set size, meaning more discriminable target-distractor pairs are required as set size increases. In signal detection theory terms, this result is consistent with any theory of attention where decisions are made with a low threshold.

For a detailed walkthrough of how we did this, please see [this notebook](./psychometric.ipynb).
